---
title: "Midterm"
author: "Srishti Patil"
date: "2024-11-04"
output: html_document
---

Import Data
• Correctly import the data.
```{r}
library(ggplot2)
library(fpp)
library(fpp2)
library(forecast)

#importing data
library(readr)
salesData <- read_csv("C:/Users/Srishti/Desktop/FALL24/BF/Midterm/sales.csv")

# Adding Headers to all the columns in the Data File
colnames(salesData) <- c("date", "sales")
head(salesData)
```


Plot and Inference
• Show a time series plot.
• Please summarize your observations of the time series plot


```{r}
salesTimeSeries <- ts(salesData$sales, frequency = 12, start = c(2020,1))
plot(salesTimeSeries)

```


Summary of the plot:
-In the given plot we can see that there's significant increase in the sales over the years from 2020 to 2024.
-Plot shows that the sales start is low level in 2020 and rise gradually and evidently fluctuate from 2023 suggesting seasonality and cyclical patterns.
-Sales starts declining from around 2024.



Central Tendency
• What are the min, max, mean, median, 1st, and 3rd Quartile values of the times series?
• Show the box plot.
• Can you summarize your observation about the time series from the summary stats and
box plot?

```{r}
summary(salesTimeSeries)
boxplot(salesTimeSeries)

```


Min: 1322
1st Quartile: 3005
Median: 7540
Mean: 8885
3rd Quartile: 14419
Max: 22397

-As shown by the boxplot, the sales varies significantly from 1322 to 22397. 
-The median being lower than the mean suggests a right-skewed distribution showing there are some high sales values pulling the mean upwards
-The IQR is 11414 which shows a wide spread in the middle 50% of the data.




Decomposition
• Plot the decomposition of the time series.

```{r}
#Using STL decomposition
decomposition_stl <- stl(salesTimeSeries,s.window ="periodic")
plot(decomposition_stl, main = 'Decomposition')
```

• Is the time series seasonal?
```{r}

#seasonal component
plot(decomposition_stl$time.series[, "seasonal"], main="Seasonal Component")
```
YES, looking at the repeating patterns in seasonal component we can see that the time series is seasonal.



• Is the decomposition additive or multiplicative?
```{r}

decompose <- decompose(salesTimeSeries)
decompose$type
```

The consistancy in the scale of seasonal and reminder components suggests that the decompositon is additive.


• If seasonal, what are the values of the seasonal monthly indices?
```{r}

decompose$figure

```


• For which month is the time series value high, and for which month is it low?

- May has the highest value
- December has the lowest value


• Can you think of the reason behind the high and low values in those months?

-High demand in May can be due to seasonal demand.
-It can also by because of increased healthcare campaigns and promotional activities in summers
-Lowest value in December can be because of the holiday seasons.
- Also, as insurance policies or benefits can often reset at the start of the tear hence people might hold on to their December purchase till January.


• Show the plot for time series adjusted for seasonality. Overlay this with the line for
actuals? Does seasonality have big fluctuations in the value of time series?

```{r}
plot(salesTimeSeries)
lines(seasadj(decomposition_stl), main = "Plot of seasonal adj", col="Blue")
```
The seasonality have big fluctuations in the value of time series. The differences in peaks and dips in the seasonally adjusted series and original series shows that due to seasonal effect the variation happened significantly which is smoothed in the blue line after we do adjustments in seasonality.



Naïve Method
• Output
```{r}
naiveMethodSales = naive(salesTimeSeries)
plot(naiveMethodSales, main = "Naive Forecast")
attributes(naiveMethodSales)
```
• Perform Residual Analysis for this technique.
o Do a plot of residuals. What does the plot indicate?
o Do a Histogram plot of residuals. What does the plot indicate?
o Do a plot of fitted values vs. residuals. What does the plot indicate?
o Do a plot of actual values vs. residuals. What does the plot indicate?
o Do an ACF plot of the residuals? What does this plot indicate?

```{r}
#Plot of residual
naiveResidual <- naiveMethodSales$residuals
plot(naiveResidual, main="Naive residual plot")

#Histogram plot of residuals
hist(naiveResidual, main="Histogram of residual plot")

#plot of fitted values vs. residuals
naiveFitted <- naiveMethodSales$fitted
plot(naiveFitted, naiveResidual, xy.labels = FALSE,xy.lines = FALSE)

#plot of actual values vs. residuals

plot(salesTimeSeries, naiveResidual,  xy.labels = FALSE,xy.lines = FALSE)

#ACF plot of the residuals
Acf(naiveResidual)

# measures of accuracy
naiveAccuracy<-accuracy(naiveMethodSales)
naiveAccuracy

#forecast
naiveforecast <- forecast(naiveMethodSales)
naiveforecast
accuracy(naiveforecast)
plot(naiveforecast)
```

Residual analysis:
-From around 2022 onward, the residuals start to show larger fluctuations, both positive and negative. Alos, the naive model's residuals have increasing variability and a clear seasonal pattern, especially from 2023 onward.
-This suggests that the naive model has increasing difficulty accurately capturing the actual values as time progresses, possibly due to rising seasonality or trend in the data.
-This histogram of the residuals shows a slight positive skew, with most residuals clustering around zero but with some higher positive values.The lack of a normal, symmetric distribution suggests the model may not fully capture all patterns in the data.
- fitted vs residual graph: This graph pattern suggests that the naive model may be systematically underestimating high values and overestimating low values, indicating potential model bias.
- actal vs residual: this graph shows a clear downward trend, indicating that as the actual sales increase, the residuals tend to become more negative.




• Print the five measures of accuracy for this forecasting technique
• Forecast
4
o Time series value for next year. Show table and plot
• Summarize this forecasting technique
o How good is the accuracy?


o What does it predict the time series value will be in one year?
Remain steady with a Point Forecast of 10,151.93 for each month




o Other observation
-The model shows moderate accuracy, with reasonable but not exceptional forecasting performance. It captures the trend to some extent but could be improved, particularly for handling high-value fluctuations and residual patterns.



Simple Moving Averages
• Plot the graph for the time series.
• Show the Simple Moving average of order three on the plot above in Red
• Show the Simple Moving average of order six on the plot above in Blue
• Show the Simple Moving average of order nine on the plot above in Green
• (Bonus) show the forecast for the next 12 months using one of the simple average
orders that you feel works best for time series
• What are your observations of the plot as the moving average order goes up?

```{r}
#Moving Avg
MAforecast1 = ma(salesTimeSeries,order=3)
MAforecast2 = ma(salesTimeSeries,order=6)
MAforecast3 = ma(salesTimeSeries,order=9)

plot(salesTimeSeries)
lines(MAforecast1, col="red")
lines(MAforecast2, col="blue")
lines(MAforecast3, col="green")

MAforecast <- ma(salesTimeSeries,12)
plot(MAforecast)
lines(MAforecast, col="purple")

```

This follows the upward trend from 2020 until early 2023, including seasonal variability especially after 2022. On the end of the year 2024, sales show a significant drop. The colored lines represent the various forecasting methods when the order is set to 3,6,and 9. Each closely follows the actual data through late 2023, where they started to diverge in different directions, indicative of uncertainty about future projections.




Simple Smoothing
• Perform a simple smoothing forecast for the next 12 months for the time series.
o What is the value of alpha? What does that value signify?

    alpha = 0.432 
    Alpha specifies the coefficient for the level smoothing. Values near 1.0 mean that the latest value has more weight.

o What is the value of the initial state?
l = 1446.3503

o What is the value of sigma? What does the sigma signify?
sigma:  2627.533
Sigma defines the variance in the forecast predicted.


• Perform Residual Analysis for this technique.
o Do a plot of residuals. What does the plot indicate?
o Do a Histogram plot of residuals. What does the plot indicate?
o Do a plot of fitted values vs. residuals. What does the plot indicate?
o Do a plot of actual values vs. residuals. What does the plot indicate?
o Do an ACF plot of the residuals? What does this plot indicate?
• Print the five measures of accuracy for this forecasting technique
• Forecast
o Time series value for next year. Show table and plot
• Summarize this forecasting technique
o How good is the accuracy?
o What does it predict the time series value will be in one year?
o Other observation
```{r}
#simple smoothing
simplesmoothing <- ses(salesTimeSeries)
plot(simplesmoothing)
summary(simplesmoothing)
attributes(simplesmoothing)

#accuracy
accuracy(simplesmoothing)
simplesmoothingaccuracy <- ses(salesTimeSeries,12)


simplesmoothingaccuracy
plot(simplesmoothingaccuracy)


plot(simplesmoothingaccuracy$residuals)
hist(simplesmoothingaccuracy$residuals)
Acf(simplesmoothingaccuracy$residuals)
forecast(simplesmoothingaccuracy)

```



Holt-Winters
• Perform Holt-Winters forecast for the next 12 months for the time series.
o What is the value of alpha? What does that value signify?

    alpha = 0.2948 
   
Alpha specifies the coefficient for the level smoothing in Holtwinters.


o What is the value of beta? What does that value signify?
 beta  = 0.0894 
    

Beta specifies the coefficient for the trend smoothing in Holtwinters.

o What is the value of gamma? What does that value signify?
gamma = 1e-04 

  

Gamma specifies the coefficient for the seasonal smoothing in Holtwinters.



o What is the value of initial states for the level, trend, and seasonality? What do
these values signify?
Initial states:
    l = 446.9245 
    b = 354.2206 
    s = -1661.38 1750.277 -1588.956 -1283.547 1035.267 -838.3166
           1015.053 2130.941 124.725 -81.034 1045.168 -1648.196


o What is the value of sigma? What does the sigma signify?
sigma:  2420.948
Sigma defines the variance of the forecast values.


• Perform Residual Analysis for this technique.
o Do a plot of residuals. What does the plot indicate?
o Do a Histogram plot of residuals. What does the plot indicate?
o Do a plot of fitted values vs. residuals. What does the plot indicate?
o Do a plot of actual values vs. residuals. What does the plot indicate?
o Do an ACF plot of the residuals? What does this plot indicate?
• Print the five measures of accuracy for this forecasting technique
• Forecast
5
o Time series value for next year. Show table and plot
• Summarize this forecasting technique
o How good is the accuracy?
The model's accuracy shows moderate error levels, with a Mean Absolute Percentage Error indicating that forecasts deviate from average from actual values. The low Autocorrelation suggests minimal correlation in residuals, which is a positive sign for forecast reliability, though the high RMSE  points to notable prediction errors.


o What does it predict the time series value will be in one year?
o Other observation


```{r}
holtsWinter <- hw(salesTimeSeries, seasonal = "additive")
plot(holtsWinter)
holtsWinter$model

summary(holtsWinter)
holtsWinter1 <- hw(salesTimeSeries,12)
plot(holtsWinter)
print(holtsWinter)

hwaccuracy<- accuracy(holtsWinter1)
hwaccuracy
plot(holtsWinter1$residuals)
hist(holtsWinter1$residuals)
Acf(holtsWinter1$residuals)
forecast(holtsWinter1)
```
Observations:
The forecast shows a general decline in sales from August 2024 to May 2025, with increasing uncertainty over time, as indicated by the widening confidence intervals. Negative lower bounds in early 2025 highlight high forecast volatility and potential model limitations for long-term predictions.




Accuracy Summary
• Show a table of all the forecast methods above with their accuracy measures.
• Separately define each forecast method and why it is useful. Show the best and worst
forecast method for each of the accuracy measures.



Conclusion
• Summarize your analysis of time series value over the time period.


• Based on your analysis and forecast above, do you think the value of the time series will
increase, decrease, or stay flat over the next year? How about the next 2 years?
according to my analysis the forecast will decrease. evenafter 2 years the forecast will decrease.

• Rank forecasting methods for this time series based on historical values.
The best forecasting method is holtswinter followed by simple smoothing then naive




